---
title: "BacTools: Discovering Bacterial Bioinformatics Tools Across CRAN, Bioconductor, and GitHub"
author: "Tasmine"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{BacTools: Discovering Bacterial Bioinformatics Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # set to TRUE when running live; FALSE keeps output static for CRAN checks
)
```

## Introduction

Bacterial bioinformatics is a rapidly growing field, with tools scattered across
CRAN, Bioconductor, and GitHub. Finding the right tool for a task — assembly,
resistance profiling, microbiome analysis — often requires searching each
repository separately.

**BacTools** solves this by providing a unified search interface that queries all
three repositories simultaneously, returning a single tidy catalog you can filter,
explore, and export.

### What BacTools does

1. Searches **CRAN** and **Bioconductor** by keyword tags in package titles and descriptions
2. Searches **GitHub** by repository topics
3. Combines results into a single data frame with consistent columns
4. Optionally retrieves DOIs and citation counts
5. Exports the catalog to CSV for downstream use

---

## Installation

```{r install, eval=FALSE}
# Install from GitHub (development version)
# install.packages("remotes")
remotes::install_github("yourusername/BacTools")
```

```{r load}
library(BacTools)
```

---

## Example Data

For this tutorial we use the package's built-in default search terms so no
external files are required. The default tags reflect common
bacterial bioinformatics keywords:

```{r show-defaults}
# Tags used to search CRAN, Bioconductor, and GitHub
BacTools::default_bacterial_tags
```

---

## Happy Path: Full Catalog Search

The main function is `search_bactools()`. Below we run it on CRAN and
Bioconductor (the example data), skipping GitHub so no token is needed for
the demo.

```{r run-search}
catalog <- search_bactools(
  use_defaults   = TRUE,
  include_cran   = TRUE,
  include_bioc   = TRUE,
  include_github = FALSE   # set TRUE and supply github_token for full run
)
```

The function prints a live summary as it runs:

```
============================================================
Querying CRAN...
============================================================
Searching CRAN with 9 tags: bacteria, bacteriology, microbiome, ...
Found 172 CRAN packages

============================================================
Querying Bioconductor...
============================================================
Found 116 Bioconductor packages

============================================================
SUMMARY: Found 172 CRAN, 116 Bioconductor, 0 GitHub tools (Total: 288 unique)
============================================================
```

---

## Inspecting the Output

The result is a plain data frame — easy to explore with base R or tidyverse tools.

```{r inspect}
# Dimensions
dim(catalog)

# Column names
colnames(catalog)

# First few rows
head(catalog[, c("name", "source", "summary", "license", "release_date")])
```

### Columns returned

| Column        | Description                                      |
|---------------|--------------------------------------------------|
| `name`        | Package / repository name                        |
| `author`      | Author(s)                                        |
| `summary`     | One-line description                             |
| `doi`         | DOI (where available)                            |
| `apa_citation`| APA-formatted citation string                    |
| `language`    | Programming language                             |
| `release_date`| Date of first / latest release                   |
| `docs`        | Link to documentation                            |
| `tutorials`   | Link to vignettes / tutorials                    |
| `license`     | Software license                                 |
| `github_stars`| GitHub stars (GitHub results only)               |
| `source`      | Repository source: CRAN, Bioconductor, or GitHub |
| `category`    | Auto-assigned tool category                      |
| `catalog_date`| Timestamp of this search                         |
| `owner`       | GitHub repository owner (GitHub results only)    |

---

## Filtering the Catalog

Because the output is a standard data frame you can filter it however you like.

```{r filter}
library(dplyr)

# Tools related to antimicrobial resistance
amr_tools <- catalog |>
  filter(grepl("resist|AMR|antibiotic", summary, ignore.case = TRUE))

nrow(amr_tools)
head(amr_tools[, c("name", "source", "summary")])

# Only Bioconductor tools
bioc_tools <- catalog |> filter(source == "Bioconductor")

# Tools released after 2022
recent_tools <- catalog |>
  filter(as.Date(substr(release_date, 1, 10)) > as.Date("2022-01-01"))

nrow(recent_tools)
```

---

## Adding Custom Search Terms

You can extend the default search with your own keywords without replacing them.

```{r custom-tags}
custom_catalog <- search_bactools(
  use_defaults    = TRUE,
  additional_tags = c("plasmid", "phage", "virulence"),
  include_cran    = TRUE,
  include_bioc    = TRUE,
  include_github  = FALSE
)

nrow(custom_catalog)
```

Or search with entirely custom terms:

```{r custom-only}
assembly_tools <- search_bactools(
  tags           = c("assembly", "assembler", "contig"),
  use_defaults   = FALSE,
  include_cran   = TRUE,
  include_bioc   = TRUE,
  include_github = FALSE
)
```

---

## Including GitHub

To include GitHub repositories, supply a personal access token. Create one at
<https://github.com/settings/tokens> (no special scopes needed for public repos).

```{r github, eval=FALSE}
full_catalog <- search_bactools(
  use_defaults   = TRUE,
  include_cran   = TRUE,
  include_bioc   = TRUE,
  include_github = TRUE,
  github_token   = Sys.getenv("GITHUB_PAT"),  # store token in .Renviron
  github_per_page = 100,
  github_pages    = 5
)

# GitHub-specific columns
full_catalog |>
  filter(source == "GitHub") |>
  arrange(desc(github_stars)) |>
  head(10) |>
  select(name, summary, github_stars, docs)
```

---

## Exporting the Catalog

Save the catalog to a timestamped CSV for sharing or downstream analysis.

```{r export}
export_catalog(catalog, filename = "bacterial_tools_catalog.csv")
# Saves as: bacterial_tools_catalog_20260325_120000.csv

# Without timestamp
export_catalog(catalog,
               filename          = "bacterial_tools_catalog.csv",
               include_timestamp = FALSE)
```

---

## Summary

In a few lines of code, BacTools:

1. **Searched** CRAN and Bioconductor for bacterial bioinformatics packages
2. **Combined** results into a single tidy data frame with 288 unique tools
3. **Enabled** easy filtering by source, keyword, date, or license
4. **Exported** the catalog to CSV

The same workflow scales to GitHub and accepts any custom search terms,
making BacTools a flexible starting point for systematic tool discovery in
bacterial bioinformatics.

---

## Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
